\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage[hidelinks]{hyperref}
\usepackage[capitalize]{cleveref}

\addbibresource{references.bib}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{construction}[theorem]{Construction}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\hash}{\mathsf{H}}
\newcommand{\digest}{\mathsf{d}}
\newcommand{\bits}{\{0,1\}}
\newcommand{\secpar}{\lambda}
\newcommand{\adv}{\mathcal{A}}
\newcommand{\negl}{\mathsf{negl}}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible,
}

\title{Authenticated Data Structures: A Merkle-Tree Log with
  Logarithmic Membership Proofs}
\author{Danny Willems}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
An authenticated data structure (ADS) lets an untrusted server answer
queries over a dataset while returning a short proof that a client can
check against a small digest of the data. This note formalises an ADS
for an append-only log built from a binary Merkle tree over a
collision-resistant hash function. We state the completeness and
soundness properties, give the tree construction and the membership
proof algorithm, and show that a membership proof for a log of $n$
entries has size $\lceil \log_2 n \rceil$ hashes and is verified with
the same number of hash evaluations. A reference implementation in
Python accompanies the construction.
\end{abstract}

\section{Introduction}
\label{sec:intro}

A client with limited storage delegates a dataset to a server and later
queries it. The client keeps a short \emph{digest} of the data; the
server answers each query with a result and a \emph{proof}. An
authenticated data structure (ADS) is the pair of algorithms that
produces such proofs and checks them against the digest, so that a
server that returns a wrong answer is detected. The model was studied
under the names memory checking~\cite{BEGKN94} and authenticated data
structures~\cite{Tamassia03}; Merkle trees~\cite{Merkle88} are the
standard building block.

This note covers one concrete ADS: an append-only log whose digest is
the root of a binary Merkle tree over the log entries. The construction
supports two queries, appending an entry and proving that a given entry
sits at a given position. We state the properties the ADS is intended to
satisfy, give the algorithms, and record the proof sizes. The accompanying
code implements the tree, the prover, and the verifier.

\paragraph{Contribution.}
We (i) state completeness and soundness for the append-only log ADS
(\cref{sec:model}), (ii) give the Merkle-tree construction and the
membership proof and verification algorithms (\cref{sec:construction}),
and (iii) bound the proof size and the verification cost
(\cref{sec:analysis}). The construction is not new; the aim is a
self-contained statement paired with a tested reference implementation.

\paragraph{Scope.}
We treat only membership (inclusion) proofs for a static or append-only
log. Non-membership proofs, range queries, and dynamic dictionaries with
updates are out of scope and are noted as future work
(\cref{sec:conclusion}).

\section{Preliminaries}
\label{sec:prelim}

We write $\secpar$ for the security parameter and $\bits^{*}$ for the set
of finite binary strings. A function $\negl(\secpar)$ is negligible if it
is smaller than every inverse polynomial for large enough $\secpar$.

\begin{definition}[Collision resistance]
\label{def:cr}
A hash function $\hash \colon \bits^{*} \to \bits^{\secpar}$ is
collision resistant if for every probabilistic polynomial-time adversary
$\adv$ the probability, over the choice of $\hash$ from the family, that
$\adv$ outputs $x \neq x'$ with $\hash(x) = \hash(x')$ is
$\negl(\secpar)$.
\end{definition}

We model the hash function as a fixed collision-resistant function and
suppress the family index for readability. Concatenation is written
$x \mathbin{\|} y$, and we assume a prefix-free encoding of the arguments
so that the concatenation is unambiguous.

\section{Model}
\label{sec:model}

An authenticated append-only log is a tuple of algorithms
$(\mathsf{Digest}, \mathsf{Prove}, \mathsf{Verify})$ over a sequence of
entries $L = (e_0, \dots, e_{n-1})$ with $e_i \in \bits^{*}$.

\begin{itemize}
  \item $\mathsf{Digest}(L) \to \digest$ outputs a digest in
    $\bits^{\secpar}$.
  \item $\mathsf{Prove}(L, i) \to \pi$ outputs a membership proof for the
    entry at position $i$.
  \item $\mathsf{Verify}(\digest, i, e, \pi) \to \bits$ accepts or rejects
    the claim that the entry at position $i$ of a log with digest
    $\digest$ equals $e$.
\end{itemize}

\begin{definition}[Completeness]
\label{def:completeness}
The ADS is complete if for every log $L$ of length $n$ and every index
$i < n$,
\[
  \mathsf{Verify}\bigl(\mathsf{Digest}(L),\, i,\, e_i,\,
  \mathsf{Prove}(L, i)\bigr) = 1 .
\]
\end{definition}

\begin{definition}[Soundness]
\label{def:soundness}
The ADS is sound if for every probabilistic polynomial-time adversary
$\adv$ the probability that $\adv$ outputs $(L, i, e, \pi)$ with
$e \neq e_i$ and
$\mathsf{Verify}(\mathsf{Digest}(L), i, e, \pi) = 1$ is
$\negl(\secpar)$.
\end{definition}

Soundness says that a proof accepted against the honest digest pins down
the entry at each position: the server cannot open position $i$ to a
value other than the one that produced the digest.

\section{Construction}
\label{sec:construction}

\begin{construction}[Merkle-tree log]
\label{con:merkle}
Fix a collision-resistant $\hash$. Let $L = (e_0, \dots, e_{n-1})$.
Define leaf digests $v_i = \hash(\texttt{0x00} \mathbin{\|} e_i)$ and
internal digests $v = \hash(\texttt{0x01} \mathbin{\|} v_{\mathrm{left}}
\mathbin{\|} v_{\mathrm{right}})$. The tree is the balanced binary tree
over the $n$ leaves; when a level has an odd number of nodes the last
node is promoted unchanged to the next level. The digest
$\mathsf{Digest}(L)$ is the root.
\end{construction}

The domain separation tags \texttt{0x00} and \texttt{0x01} distinguish
leaf hashing from internal hashing, which blocks the second-preimage
attack that identifies a leaf with an internal node.

$\mathsf{Prove}(L, i)$ returns the sibling digest at each level on the
path from leaf $i$ to the root, together with a bit per level recording
whether the sibling is on the left or the right. $\mathsf{Verify}$
recomputes the path: it hashes $e$ into a leaf digest, folds in each
sibling according to its side bit, and accepts iff the recomputed root
equals $\digest$. \Cref{alg:verify} gives the verifier.

\begin{algorithm}
\caption{$\mathsf{Verify}(\digest, i, e, \pi)$}
\label{alg:verify}
\begin{algorithmic}[1]
  \State $v \gets \hash(\texttt{0x00} \mathbin{\|} e)$
  \For{each $(s, \mathit{side})$ in $\pi$ from leaf to root}
    \If{$\mathit{side} = \textsc{left}$}
      \State $v \gets \hash(\texttt{0x01} \mathbin{\|} s \mathbin{\|} v)$
    \Else
      \State $v \gets \hash(\texttt{0x01} \mathbin{\|} v \mathbin{\|} s)$
    \EndIf
  \EndFor
  \State \Return $v = \digest$
\end{algorithmic}
\end{algorithm}

\section{Analysis}
\label{sec:analysis}

\begin{proposition}[Completeness]
\label{prop:completeness}
\Cref{con:merkle} is complete in the sense of
\cref{def:completeness}.
\end{proposition}

\begin{proof}
$\mathsf{Prove}(L, i)$ returns exactly the sibling digests along the
leaf-to-root path, in order, with correct side bits. The fold in
\cref{alg:verify} applies the same internal hashing as
\cref{con:merkle} at each level, in the same order, so the recomputed
value equals the honest root, which is $\digest$. Hence $\mathsf{Verify}$
returns $1$.
\end{proof}

\begin{proposition}[Soundness]
\label{prop:soundness}
If $\hash$ is collision resistant (\cref{def:cr}), then
\cref{con:merkle} is sound in the sense of \cref{def:soundness}.
\end{proposition}

\begin{proof}[Proof sketch]
Suppose an adversary produces $(L, i, e, \pi)$ with $e \neq e_i$ that
verifies against $\digest = \mathsf{Digest}(L)$. The honest path from
leaf $i$ and the path recomputed by the verifier both start at the root
$\digest$ and both are folds of $\hash$. Since $e \neq e_i$, the two
paths differ at the leaf. Walk down from the root to the first level
where the two paths hold different values whose parent holds the same
value: at that level two distinct hash inputs map to the same digest,
which is a collision for $\hash$. Extracting this collision runs in
polynomial time, so by \cref{def:cr} the adversary's success probability
is $\negl(\secpar)$.
\end{proof}

\begin{proposition}[Proof size and verification cost]
\label{prop:size}
For a log of $n$ entries a membership proof consists of
$\lceil \log_2 n \rceil$ sibling digests, and $\mathsf{Verify}$ performs
$\lceil \log_2 n \rceil + 1$ evaluations of $\hash$.
\end{proposition}

\begin{proof}
The balanced tree over $n$ leaves has height $\lceil \log_2 n \rceil$,
and the path from a leaf to the root visits one sibling per level. The
verifier hashes the claimed entry into a leaf digest (one evaluation)
and then performs one internal evaluation per level.
\end{proof}

\section{Reference implementation}
\label{sec:impl}

The accompanying Python package \texttt{ads} implements
\cref{con:merkle}. The module \texttt{ads.merkle} builds the tree,
produces membership proofs, and verifies them with the domain-separated
hashing of \cref{con:merkle}; the default hash is SHA-256. The test
suite checks completeness on random logs and checks that tampering with
an entry, an index, or a proof causes verification to reject. Proof
sizes measured in the tests match \cref{prop:size}.

A companion Lean~4 development (directory \texttt{lean/}) machine-checks
the three properties over an abstract leaf-hash and node-hash, the two
distinct functions modelling the domain separation of \cref{con:merkle}.
The theorems \texttt{Ads.completeness}, \texttt{Ads.soundness}, and
\texttt{Ads.proof\_length} correspond to \cref{prop:completeness},
\cref{prop:soundness}, and \cref{prop:size}; \texttt{Ads.soundness}
states the contrapositive of \cref{def:soundness}, that a wrong opening
yields a hash collision. The proofs use no \texttt{sorry} and depend only
on the standard axioms \texttt{propext}, \texttt{Classical.choice}, and
\texttt{Quot.sound}.

\section{Conclusion}
\label{sec:conclusion}

We stated an authenticated append-only log ADS from a binary Merkle
tree, with logarithmic proof size and verification cost, and paired it
with a tested implementation. Directions not covered here include
non-membership proofs over a sorted structure, consistency proofs
between two digests of the same growing log (as in certificate
transparency), and authenticated dictionaries supporting updates.

\printbibliography

\end{document}
